Oregon GWR smoke association with Cardiopulmonary Emergency Room visits

### 2018-05-30 r_scripts/analysis/oregon_wfs_ed_results.Rmd
renamed part_a and removed some code. Changed figure colors and added back in arrhythmia.

---
title: "Oregon GWR smoke association with Cardiopulmonary Emergency Room visits (PartA)"
author: "Jingyang Liu"
date: "April 9, 2018"
output: html_document
---

@@

This project researches on for the association between GWR smoke PM~2.5~ and cardiopulmonary (cardiovascular (CVD) and respiratory health outcomes) diseases for Oregon State in wildfire season (May to September) in 2013. 
This markdown document focuses on the zip code-level analysis. 
The analysis includes total analysis, age strata and sex strata.

@@

In these research, we mainly examine GWR smoke/PM~2.5~ estimations and associations with health outcomes using a time-stratified case-crossover study design. Health outcomes for a patient with a primary diagnosis of cardiopulmonary health outcomes and their date of admission  (index time) were identified. We then created counterfactual observations for each patient for the same day of the week for the entire wildfire season (MAy 1st to Sep 30th, 2013). We further limited our analyses to claims that were coded as emergency or urgent visits to eliminate bias from patients going in for elective/planned procedures. However, we found the filtered population is much larger than the population we chose in Washington, so we choose only the emegency room (hospital) visits and urgent care visits. For patients who have a index time closer to May 1, 2013 or closer to September 30, 2013, their referent observations before or after these dates will be excluded as I will not be able to assign estimates of PM~2.5~ to these referent observations. However, this is not a big issue for a time-stratified design as other referent values throughout the time period average out.

The **cardiopulmonary health outcomes** are all respiratory, asthma, COPD, cardiovascular disease, heart failure, and ischemic heart disease.

As for **exposure methods**, we use Geo-Weighted Regression and subtracted off the 'Background' estimates of smoke, which I believe are the monthly averages of PM~2.5~ for a given grid. However, the 'Background' is the mean of total period from May 1st to Sep 30th 2013 so I have to assign the mean for every day. Then use the 10 units of the smoke for analysis.

The **analytic method** is the conditional logistic regression model using the *survival* package in R. This regression is controlling the confounders. Each conditional logistic regression model accounts for the subject, and adjusts for temperature from the WRF-Chem model. The conditional logistic regression model \beta represents the change in risk of an event associated with a short-term 10 units' increase in exposure, and can be calculated as an average difference between exposure at the index time and a weighted average of exposure at all times.

**Library:**  
tidyverse is for data cleaning, also includes ggplot and readr.  
data.table is for quick reading files.  
htmlTable is for writing table.  
survival is for running conditional logistic regression model.

@@

for (i in var_list){
  read_path <- paste('../data_new/casecrossover/time_strat/or', i, 'may_to_sep_time_strat_casecross.csv', sep = "_")
  or_disease <- read_csv(read_path)

@@

```{r load ER data set and zip smoke, echo=FALSE, message=FALSE, warning=FALSE}
## Analysis
path <- paste0("../data_new/casecrossover/ER/")

@@

# pneum or bronchitis
pneum_casecross <- read_csv(paste(path, 
                                  "or_pneumonia_may_to_sep_time_strat_casecross_er.csv",
                                  sep="/"))
 # acute bronchitis
acute_bronch_casecross <- read_csv(paste(path, 
                                         "or_acute_bronch_may_to_sep_time_strat_casecross_er.csv",
                                         sep = "/"))
# cvd
cvd_casecross <- read_csv(paste(path, 
                                "or_cvd_may_to_sep_time_strat_casecross_er.csv",
                                sep="/"))
# arrhythmia
arrhythmia_casecross <- read_csv(paste(path, 
                                       "or_arrhythmia_may_to_sep_time_strat_casecross_er.csv",
                                       sep="/"))
# cerebral vascular
cereb_vas_casecross <- read_csv(paste(path, 
                                      "or_cerbrovas_dis_may_to_sep_time_strat_casecross_er.csv",
                                      sep="/"))
# heart failure
hf_casecross <- read_csv(paste(path, 
                               "or_heart_failure_may_to_sep_time_strat_casecross_er.csv", 
                               sep="/"))
# ischemic heart disease
ihd_casecross <- read_csv(paste(path, 
                                "or_isch_heart_dis_may_to_sep_time_strat_casecross_er.csv",
                                sep="/"))
# myo infarc
mi_casecross <- read_csv(paste(path, 
                               "or_myocardial_infarc_may_to_sep_time_strat_casecross_er.csv", sep="/"))

@@

# dataframe list
df_list <- list(resp_casecross, asthma_casecross, copd_casecross, 
  pneum_casecross, acute_bronch_casecross, cvd_casecross, arrhythmia_casecross,
  cereb_vas_casecross, hf_casecross, ihd_casecross, mi_casecross)

outcome_list <- c('All Respiratory', 'Asthma', 'COPD', 
                  'Pneumonia', 'Acute Bronchitis', 'Cardiovascular Disease',
                  'Arrhythmia', 'Cerebrovascular Disease', 'Heart Failure',
                  'Ischemic Heart Disease', 'Myocardial Infarction')

@@

# combine each outcome dataframe itteration in to a big dataset
combined_point_est_df <- bind_rows(datalist) %>% 
  # find proportions/percents for each strata in a row
  mutate(age_15_pr = as.character(round((as.numeric(age_15)/as.numeric(n))*100,1)),
         age_15_65_pr = as.character(round((as.numeric(age_15_65)/as.numeric(n))*100,1)),
         age_65_pr = as.character(round((as.numeric(age_65)/as.numeric(n))*100,1)),
         female_pr = as.character(round((as.numeric(female)/as.numeric(n))*100,1)),
         male_pr = as.character(round((as.numeric(male)/as.numeric(n))*100,1))) %>% 
  select(outcome, n, age_15, age_15_pr, age_15_65, age_15_65_pr, age_65,
         age_65_pr, female, female_pr, male, male_pr)

#str(combined_point_est_df)

tab <- htmlTable(txtRound(combined_point_est_df, digits = 1), 
           caption = "Number of cases for each outcome observed from May 1st to September 30st, 2013",
           # column headers
           header = c("Outcome", "Cases n", "Less than 15", "(%)", "15 to 65", "(%)", "Greater than 65",
                      "(%)", "Female", "(%)", "Male", "(%)"),
           # column spanner
           cgroup = c("","Age Category", "Sex"), 
           n.cgroup = c(2, 6, 4),
           padding.rgroup = "&nbsp;&nbsp;",
           css.cell = "padding-left: 0.5em; padding-right: .5em;", # cell space
           align = "llccccc" # column alignment,
            ) # end table

print(tab)

@@

```{r echo = F, warning = F, results='asis'}

df_list <- list(resp_casecross, asthma_casecross, copd_casecross, 
                pneum_casecross, acute_bronch_casecross, cvd_casecross, arrhythmia_casecross,
                cereb_vas_casecross, hf_casecross, ihd_casecross, mi_casecross)

outcome_list <- c('All Respiratory', 'Asthma', 'COPD', 
                  'Pneumonia', 'Acute Bronchitis', 'Cardiovascular Disease',
                  'Arrhythmia', 'Cerebrovascular Disease', 'Heart Failure',
                  'Ischemic Heart Disease', 'Myocardial Infarction')

@@

  # dataframe for analysis creation
  # bind columns back together 
  df_analysis <- covariates_df %>% 
    # remove missing pm values
    filter(!is.na(geo_smk_pm_zip10)) %>%

    # join in cdc met data
    # full_join(daily_zip_met, by = c("ZIPCODE", "date")) %>% 
    # the following code makes sure that the counterfactual values retained are 
    # symetric in that number of obs before = number of obs after
    mutate(obs_diff_admission = (fromdate - date)/7) 
  # dataframe is already for the entire fire season, so I don't need to subset anymore

@@

    # some model/DAG checks of data rish provided

    #glimpse(df_analysis)

    # adjusting for admit month is not possible as it's the same var for the subject.
    # adjusting for month smoke is a collider and should not be adjusted for.
    #summary(clogit(outcome ~ geo_smk_pm_zip + daily_meanRH + daily_meanT + daily_meanWS + daily_meanPrcp +
    #                as.factor(month_smk) + strata(PATIENTID), df_analysis))
    #outocme association: daily_meanRH,

    # check for confoudning following DAG assumptions
    #summary(clogit(outcome ~ wrf_temp_zip + strata(PATIENTID), df_analysis))
    #summary(clogit(outcome ~ wrf_temp_zip + strata(PATIENTID), df_analysis))
    # smoke predictor association: daily_meanRH, 
    #summary(lmer(geo_smk_pm_zip ~ daily_meanWS + (1 | PATIENTID), df_analysis ))
    #summary(lmer(wrf_temp_zip ~ daily_meanT + (1 | PATIENTID), df_analysis ))
    #cor(df_analysis$wrf_temp_county, df_analysis$daily_meanT)

@@

# combine each outcome dataframe itteration in to a big dataset
combined_point_est_df <- bind_rows(datalist)

# write_path <- paste0('../data_new/casecrossover/geo_health_or.csv')
# write_csv(combined_point_est_df, write_path)

# subset columns I want to put in to the table
table_df <- combined_point_est_df %>% select(2, 3:7)

tab <- htmlTable(txtRound(table_df, digits = 3, 1:3), 
                 caption = "Association between a 10 ug/m^3 in PM2.5 and Health Outcomes",
                 # row group by outcome
                 rgroup = outcome_list,
                 n.rgroup = c(rep(1, 11)), # 4 rows for each method for each outcome
                 # column headers
                 header = c("Method", "Obs.", "Events",
                            "OR&dagger;", "Lower", "Upper"),
                 # column spanner
                 cgroup = c("", "95% CI"), 
                 n.cgroup = c(4, 2),
                 padding.rgroup = "&nbsp;&nbsp;",
                 css.cell = "padding-left: 0.5em; padding-right: .5em;", # cell space
                 align = "llccccc", # column alignment,
                 tfoot="&dagger; Adjusted for CDC temperature, relatively humidity, wind speed, and precipitation; accounting for subject. Time-stratified: referent periods matched to events on same day of week within May to Sep fire season."
) # end table

print(tab)

# ggplot of odds ratios, facet wrapped by outcomes -----
# convert variables from character to either numeric or factor
# factor preserves the order of the variable
combined_point_est_df$outcome <- factor(combined_point_est_df$outcome, 
                                        levels = unique(combined_point_est_df$outcome))

combined_point_est_df$pm_method <- factor(combined_point_est_df$pm_method, 
                                          levels = unique(combined_point_est_df$pm_method))

combined_point_est_df$n <- as.numeric(combined_point_est_df$n)
combined_point_est_df$n_events <- as.numeric(combined_point_est_df$n_events)
combined_point_est_df$odds_ratio <- as.numeric(combined_point_est_df$odds_ratio)
combined_point_est_df$lower95 <- as.numeric(combined_point_est_df$lower95)
combined_point_est_df$upper95 <- as.numeric(combined_point_est_df$upper95)
combined_point_est_df$se <- as.numeric(combined_point_est_df$se)
combined_point_est_df$p_val <- as.numeric(combined_point_est_df$p_val)

# subset out global method
combined_point_est_df <- combined_point_est_df

## ggplot
print_plot <- ggplot(combined_point_est_df, 
                     aes(x = pm_method, y = odds_ratio, colour = pm_method)) +
  # custom color 
  scale_color_manual(name = "Smoke-Estimation Method", 
                     values = c("red", "blue", "#32115C"),
                     guide = guide_legend(title.position = "top", title.hjust = 0.5)) +
  geom_point() + #geom_text(vjust = 0, nudge_x = 0.3) +
  geom_errorbar(aes(ymin=lower95, ymax=upper95), width = 0.2) +
  facet_wrap(~outcome, nrow = 3) +
  geom_hline(yintercept = 1, linetype=2) +
  #    ggtitle('Association Between PM2.5 \n from Wildfire Smoke on Hospitalizations') +
  ylab(expression(paste("Odds Ratio for 10µg/m"^3, " Increase in PM"[2.5]))) +
  #xlab('Smoke Estimation Method') +
  # plot theme
  theme(panel.background = element_rect(fill = 'white', colour = 'black'),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        # strip element
        strip.background = element_rect(colour=NA, fill=NA),
        panel.border = element_rect(fill = NA, color = "black"),
        # facet text size
        #strip.text = element_text(size = 8),
        # axis element
        axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.title.y = element_text(angle = 90),
        # legend elements
        legend.position = "bottom")
#legend.text = element_text(size = 8))

# print(print_plot)
# save figure
# ggsave("../plot_new/fig_both_er_ur.pdf", plot = print_plot, 
#        width = 12, height = 8, units = "in")

write_path <- paste0("./all_strata_result_table.csv")
write_csv(combined_point_est_df, write_path)

@@

```{r time stratified age strata cdc met, warning =F, echo = F, results='asis'} 
# note above, results = 'asis' needed to print htmlTables

# For age categories, I want a loop for each age category. 
# Presenting just WRF and GWR

# dataframe list
df_list <- list(resp_casecross, asthma_casecross, copd_casecross, pneum_casecross,
                acute_bronch_casecross, cvd_casecross, arrhythmia_casecross,
                cereb_vas_casecross, hf_casecross, ihd_casecross, mi_casecross)

outcome_list <- c('All Respiratory', 'Asthma', 'COPD', 'Pneumonia', 
                  'Acute Bronchitis', 'Cardiovascular Disease', 'Arrhythmia', 
                  'Cerebrovascular Disease', 'Heart Failure',
                  'Ischemic Heart Disease', 'Myocardial Infarction')

# looking only at wrf chem, kriging, and geo-weighted; taking out 'Global Smoke' method
method_list <- c('GWR Smoke')

# age category list
age_cat_list <- c(0,1,2)

@@

var_list <- c('respiratory', 'asthma', 'pneumonia',  'acute_bronch', 'copd', 
              'cvd', 'isch_heart_dis', 'arrhythmia', 'heart_failure', 
              'cerbrovas_dis', 'myocardial_infarc')

@@

  # bind rows of age category estimates together
 age_est_df <- bind_rows(datalist1)

@@

# combine each outcome dataframe itteration in to a big dataset
combined_point_est_df <- bind_rows(datalist2)

wrf_geo_age <- combined_point_est_df %>% 
  # filter columns to just geo-smk
  # filter(pm_method == "GWR Smoke" |
    #      pm_method == "WRF-Chem Smoke") %>% 
  mutate(age_cat2 = ifelse(age_cat == 0, "Less than 15", 
                    ifelse(age_cat == 1, "15-65",
                    ifelse(age_cat == 2, "Greater than 65", NA)))) %>% 
  # subset columns I want to put in to the table
  select(3, 11, 5:8) %>% 
  mutate(odds_ratio = ifelse(odds_ratio == 99, "--", odds_ratio),
         lower95 = ifelse(lower95 == 99, "--", lower95),
         upper95 = ifelse(upper95 == 99, "--", upper95))

  tab <- htmlTable(txtRound(wrf_geo_age, digits = 3, 1:3), 
           caption = "Association between a 10 ug/m^3 in PM2.5 geo smoke and health outcomes stratified by age",
           # row group by outcome
           rgroup = outcome_list,
           n.rgroup = c(rep(3, 11)), # 6 rows for each age cat for each outcome
           # column headers
           header = c("Est Method", "Age Group", "Events",
                      "OR&dagger;", "Lower", "Upper"),
           # column spanner
           cgroup = c("", "95% CI"), 
           n.cgroup = c(4, 2),
           padding.rgroup = "&nbsp;&nbsp;",
           css.cell = "padding-left: 0.5em; padding-right: .5em;", # cell space
           align = "llcccc", # column alignment,
           tfoot="&dagger; Adjusted for CDC temperature, wind speed, relative humidity, precipitation, and accounting for subject. Time-stratified: referent periods matched to events on same day of week within May to Sep fire season."
            ) # end table

  print(tab)

combined_point_est_df_new <- combined_point_est_df %>%
  mutate(n_events_new = ifelse(as.numeric(n_events)>=100, n_events,
                        ifelse(as.numeric(n_events)<100, "0", NA))) %>%
  select(-n_events) %>%
  rename(n_events = n_events_new)

# ggplot of odds ratios, facet wrapped by outcomes -----
# convert variables from character to either numeric or factor
# factor preserves the order of the variable
wrf_geo_age_plot <- combined_point_est_df_new %>% 
  # filter columns to just geo-smk
 #  filter(pm_method == "GWR Smoke" | pm_method == "WRF-Chem Smoke") %>% 
  # do not plot results with less than 15 events, create missing vals
  # in mutate
  mutate(odds_ratio = ifelse(as.numeric(n_events) < 15, 
                             99, odds_ratio),
         lower95 = ifelse(as.numeric(n_events) < 15, 
                             99, lower95),
         upper95 = ifelse(as.numeric(n_events) < 15, 
                             99, upper95),
         age_cat2 = ifelse(age_cat == 0, "<15", 
                    ifelse(age_cat == 1, "15-65",
                    ifelse(age_cat == 2, ">65", NA)))) %>% 
  # subset columns I want to put in to the table
  select(1, 3, 11, 10, 5:7) %>% 
  mutate(odds_ratio = ifelse(odds_ratio == 99, NA, odds_ratio),
         lower95 = ifelse(lower95 == 99, NA, lower95),
         upper95 = ifelse(upper95 == 99, NA, upper95))

# change characters to numeric and factor  
wrf_geo_age_plot$outcome <- factor(wrf_geo_age_plot$outcome,
                              levels = unique(wrf_geo_age_plot$outcome))

wrf_geo_age_plot$pm_method <- factor(wrf_geo_age_plot$pm_method,
                                levels = unique(wrf_geo_age_plot$pm_method))

wrf_geo_age_plot$age_cat2 <- factor(wrf_geo_age_plot$age_cat2,
                                levels = unique(wrf_geo_age_plot$age_cat2))

wrf_geo_age_plot$n_events <- as.numeric(wrf_geo_age_plot$n_events)
wrf_geo_age_plot$odds_ratio <- as.numeric(wrf_geo_age_plot$odds_ratio)
wrf_geo_age_plot$lower95 <- as.numeric(wrf_geo_age_plot$lower95)
wrf_geo_age_plot$upper95 <- as.numeric(wrf_geo_age_plot$upper95)

## ggplot 
  print_plot <- ggplot(wrf_geo_age_plot,
    aes(x = age_cat2, y = odds_ratio, color = pm_method), na.rm = T) +
    geom_point(position = position_dodge(width = 0.5), na.rm = T) + 
    geom_errorbar(aes(ymin=lower95, ymax=upper95), 
                  position = position_dodge(width = 0.5), width = 0.2, na.rm =T) +
    # custom color 
    scale_color_manual(name = "Smoke-Estimation Method", 
                       values = c("red",  "blue", "#32115C"),
                       guide = guide_legend(title.position = "top", title.hjust = 0.5)) +
    facet_wrap(~outcome, nrow = 3, scales = "free") +
    geom_hline(yintercept = 1, linetype=2) +
    #ggtitle('Association Between PM2.5 from \n Wildfire Smoke on Hospitalizations') +
    ylab(expression(paste("Odds Ratio for 10µg/m"^3, " Increase in PM"[2.5]))) +
    #ylim(0, 2) +
    xlab('Age Category ') +
    # plot theme
    theme(panel.background = element_rect(fill = 'white', colour = 'black'),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    # strip element
    strip.background = element_rect(colour=NA, fill=NA),
    panel.border = element_rect(fill = NA, color = "black"),
    # facet text size
    strip.text = element_text(size = 10),
    # axis element
    #axis.text.x = element_blank(),
    #axis.title.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.y = element_text(angle = 90),
    # legend elements
    legend.position = "bottom")
    #legend.text = element_text(size = 8))

  # print(print_plot)
  # save figure
  # ggsave("../plot_new/age_fig_three_both.pdf", plot = print_plot, 
  #      width = 12, height = 8, units = "in")

  write_path <- paste0("./age_result_table.csv")
write_csv(combined_point_est_df, write_path)

@@

```{r sex strata cdc met adj, warning = F, echo = F, results='asis'} 
# note above, results = 'asis' needed to print htmlTables

# For age categories, I want a loop for each age category.

# dataframe list
df_list <- list(resp_casecross, asthma_casecross, copd_casecross, pneum_casecross,
                acute_bronch_casecross, cvd_casecross, arrhythmia_casecross,
                cereb_vas_casecross, hf_casecross, ihd_casecross, mi_casecross)

outcome_list <- c('All Respiratory', 'Asthma', 'COPD', 'Pneumonia', 
                  'Acute Bronchitis', 'Cardiovascular Disease', 'Arrhythmia', 
                  'Cerebrovascular Disease', 'Heart Failure',
                  'Ischemic Heart Disease', 'Myocardial Infarction')

var_list <- c('respiratory', 'asthma', 'pneumonia',  'acute_bronch', 'copd', 
              'cvd', 'isch_heart_dis', 'arrhythmia', 'heart_failure', 
              'cerbrovas_dis', 'myocardial_infarc')

# looking only at wrf chem, kriging, and geo-weighted; taking out 'Global Smoke' method
method_list <- c('GWR Smoke')

# sex category list
sex_strata_list <- c(0,1)

@@

# combine each outcome dataframe itteration in to a big dataset
combined_point_est_df <- bind_rows(datalist2)

wrf_geo_sex <- combined_point_est_df %>% 
  # filter columns to just geo-smk
  # filter(pm_method == "GWR Smoke" | pm_method == "WRF-Chem Smoke") %>% 
  mutate(sex_cat = ifelse(sex == 0, "Male", 
               ifelse(sex == 1, "Female", NA))) %>% 
  # subset columns I want to put in to the table
  select(3, 11, 5:8)%>% 
  mutate(odds_ratio = ifelse(odds_ratio == 99, "--", odds_ratio),
         lower95 = ifelse(lower95 == 99, "--", lower95),
         upper95 = ifelse(upper95 == 99, "--", upper95))

  tab <- htmlTable(txtRound(wrf_geo_sex, digits = 3, 1:3), 
           caption = "Association between a 10 ug/m^3 in PM2.5 geo smoke and health outcomes stratified by sex",
           # row group by outcome
           rgroup = outcome_list,
           n.rgroup = c(rep(2, 11)), # 2 rows for each sex strata for each outcome
           # column headers
           header = c("Est Method", "Sex Strata", "Events",
                      "OR&dagger;", "Lower", "Upper"),
           # column spanner
           cgroup = c("", "95% CI"), 
           n.cgroup = c(4, 2),
           padding.rgroup = "&nbsp;&nbsp;",
           css.cell = "padding-left: 0.5em; padding-right: .5em;", # cell space
           align = "llcccc", # column alignment,
           tfoot="&dagger; Adjusted for temperature, accounting for subject. Time-stratified: referent periods matched to events on same day of week within May to Sep fire season."
            ) # end table

  print(tab)

# ggplot of odds ratios, facet wrapped by outcomes -----
# convert variables from character to either numeric or factor
# factor preserves the order of the variable
wrf_geo_sex_plot <- combined_point_est_df %>% 
  # filter columns to just geo-smk
 #  filter(pm_method == "GWR Smoke" | 
   #       pm_method == "WRF-Chem Smoke") %>% 
  mutate(sex_cat = ifelse(sex == 0, "Male", 
               ifelse(sex == 1, "Female", NA))) %>%
  # subset columns I want to put in to the table
  select(1, 3, 11, 5:8) %>% 
  mutate(odds_ratio = ifelse(odds_ratio == 99, NA, odds_ratio),
         lower95 = ifelse(lower95 == 99, NA, lower95),
         upper95 = ifelse(upper95 == 99, NA, upper95))

# change characters to numeric and factor  
wrf_geo_sex_plot$outcome <- factor(wrf_geo_sex_plot$outcome,
                              levels = unique(wrf_geo_sex_plot$outcome))

wrf_geo_sex_plot$pm_method <- factor(wrf_geo_sex_plot$pm_method,
                                levels = unique(wrf_geo_sex_plot$pm_method))

wrf_geo_sex_plot$sex_cat <- factor(wrf_geo_sex_plot$sex_cat,
                                levels = unique(wrf_geo_sex_plot$sex_cat))

wrf_geo_sex_plot$n_events <- as.numeric(wrf_geo_sex_plot$n_events)
wrf_geo_sex_plot$odds_ratio <- as.numeric(wrf_geo_sex_plot$odds_ratio)
wrf_geo_sex_plot$lower95 <- as.numeric(wrf_geo_sex_plot$lower95)
wrf_geo_sex_plot$upper95 <- as.numeric(wrf_geo_sex_plot$upper95)

## ggplot
  print_plot <- ggplot(wrf_geo_sex_plot,
    aes(x = sex_cat, y = odds_ratio, color = pm_method), na.rm = T) +
    geom_point(position = position_dodge(width = 0.5), na.rm = T) + 
    geom_errorbar(aes(ymin=lower95, ymax=upper95), 
                  position = position_dodge(width = 0.5), width = 0.2, na.rm =T) +
    # custom color 
    scale_color_manual(name = "Smoke-Estimation Method", 
                       values = c("red", "blue", "#32115C"),
                       guide = guide_legend(title.position = "top", title.hjust = 0.5)) +
    facet_wrap(~outcome, nrow = 3, scales = "free") +
    geom_hline(yintercept = 1, linetype=2) +
    #ggtitle('Association Between PM2.5 from \n Wildfire Smoke on Hospitalizations') +
    ylab(expression(paste("Odds Ratio for 10µg/m"^3, " Increase in PM"[2.5]))) +
    #ylim(0, 2) +
    xlab('Sex') +
    # plot theme
    theme(panel.background = element_rect(fill = 'white', colour = 'black'),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    # strip element
    strip.background = element_rect(colour=NA, fill=NA),
    panel.border = element_rect(fill = NA, color = "black"),
    # facet text size
    strip.text = element_text(size = 10),
    # axis element
    #axis.text.x = element_blank(),
    #axis.title.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.y = element_text(angle = 90),
    # legend elements
    legend.position = "bottom")
    #legend.text = element_text(size = 8))

  # print(print_plot)
  # save figure
  # ggsave("../plot_new/sex_fig_three_both.pdf", plot = print_plot, 
  #      width = 12, height = 8, units = "in")

write_path <- paste0("./sex_result_table.csv")
write_csv(combined_point_est_df, write_path)

@@

# print plot
ggsave("../plot_new/cardiopulmonary_er_result.tiff", plot = result_plot, units="in", width=12, height=8, dpi=400)

# ggsave("../plot_new/cardiopulmonary_er_result_table.pdf", plot = result_plot, 
#         width = 12, height = 8, units = "in", dpi = 400)

tiff('../plot_new/cardiopulmonary_er_result.tiff', units="in", width=12, height=8, res=300)

@@

We see that all exposure methods are significantly associated with an increase in asthma with Geo-Weighted Regression, those are associated with COPD. I believe this is the most justifiable design approach that would average any extreme values of PM~2.5~ across the wildfire season for the vector of referent values for a particular subject.

*Note from Jingyang 04/12/2018:*  
Ryan said he is more likely interested in the diseases shown in plot (7 in total), not all cardiopulmonary diseases, however, I still loaded all the disease data set and reserved them in result and descriptive table to show the result.

>>>>

---
title: "Oregon GWR smoke association with Cardiopulmonary Emergency Room visits"
author: "Jingyang Liu"
date: "April 9, 2018"
output: html_document
---

@@

This project researches on the association between geographically weighted regression (GWR) smoke PM~2.5~ and cardiopulmonary (cardiovascular (CVD) and respiratory health outcomes) diseases for Oregon State in wildfire season (May to September) in 2013. 
This markdown document focuses on the zip code-level analysis. 
The analysis includes total analysis, age strata and sex strata.

@@

In these research, we mainly examine GWR PM~2.5~ estimations and associations with health outcomes using a time-stratified case-crossover study design. The data set contains all claims of Oregon for total 2013 year, so we choose only the emegency room (hospital) visits for this analysis. Health outcomes for a patient with a primary diagnosis of cardiopulmonary health outcomes and their date of admission  (index time) were identified. We then created counterfactual observations for each patient for the same day of the week for the entire wildfire season (MAy 1st to Sep 30th, 2013). We further limited our analyses to claims that were coded as emergency or urgent visits to eliminate bias from patients going in for elective/planned procedures.  For patients who have a index time closer to May 1, 2013 or closer to September 30, 2013, their referent observations before or after these dates will be excluded as we will not be able to assign estimates of PM~2.5~ to these referent observations. However, this is not a big issue for a time-stratified design as other referent values throughout the time period average out.

The **cardiopulmonary health outcomes** are all respiratory, asthma, COPD, cardiovascular disease, heart failure, and ischemic heart disease and cerebrovascular disease.

As for **exposure methods**, we use Geo-Weighted Regression and subtracted off the 'Background' estimates, which are the monthly averages of PM~2.5~ for a given grid.

The **analytic method** is the conditional logistic regression model using the *survival* package in R. This regression is controlling the time-invariant subject-specific counfounding like age and sex. Each conditional logistic regression model accounts for the subject, and adjusts for temperature from the WRF-Chem model (Weather Research and Forecasting model coupled with Chemistry). The conditional logistic regression model \beta represents the change in risk of an event associated with a short-term 10 units' increase in exposure, and can be calculated as an average difference between exposure at the index time and a weighted average of exposure at all times.

**Library:**  
tidyverse is for data cleaning, also includes ggplot and readr.  
data.table is for quick reading files.  
survival is for running conditional logistic regression model.  
htmlTable is for writing table.

@@

for (i in var_list){
  read_path <- paste('../../data/health/or', i, 
                     'may_to_sep_time_strat_casecross.csv', sep = "_")
  or_disease <- read_csv(read_path)

@@

```{r load ER data set and zip smoke, echo=FALSE, message=FALSE, warning=FALSE}
## Analysis
path <- paste0("../../data/health/")

@@

# cvd
cvd_casecross <- read_csv(paste(path, 
                                "or_cvd_may_to_sep_time_strat_casecross_er.csv",
                                sep="/"))
# cerebral vascular
cereb_vas_casecross <- read_csv(paste(path, 
                                      "or_cerbrovas_dis_may_to_sep_time_strat_casecross_er.csv",
                                      sep="/"))

# arrythmia
arrhythmia_casecross <- read_csv(paste(path, 
                               "or_arrhythmia_may_to_sep_time_strat_casecross_er.csv", 
                               sep="/"))

# heart failure
hf_casecross <- read_csv(paste(path, 
                               "or_heart_failure_may_to_sep_time_strat_casecross_er.csv", 
                               sep="/"))
# ischemic heart disease
ihd_casecross <- read_csv(paste(path, 
                                "or_isch_heart_dis_may_to_sep_time_strat_casecross_er.csv",
                                sep="/"))

@@

# dataframe list
df_list <- list(resp_casecross, asthma_casecross, copd_casecross, 
  cvd_casecross, arrhythmia_casecross, cereb_vas_casecross, hf_casecross, 
  ihd_casecross)

outcome_list <- c('All Respiratory', 'Asthma', 'COPD', 
                  'Cardiovascular Disease', 'Arrhythmia', 
                  'Cerebrovascular Disease', 'Heart Failure',
                  'Ischemic Heart Disease')

@@

# combine each outcome dataframe itteration in to a big dataset
tab1 <- bind_rows(datalist) %>% 
  # find proportions/percents for each strata in a row
  mutate(age_15_pr = as.character(round((as.numeric(age_15)/as.numeric(n))*100,1)),
         age_15_65_pr = as.character(round((as.numeric(age_15_65)/as.numeric(n))*100,1)),
         age_65_pr = as.character(round((as.numeric(age_65)/as.numeric(n))*100,1)),
         female_pr = as.character(round((as.numeric(female)/as.numeric(n))*100,1)),
         male_pr = as.character(round((as.numeric(male)/as.numeric(n))*100,1))) %>% 
  select(outcome, n, age_15, age_15_pr, age_15_65, age_15_65_pr, age_65,
         age_65_pr, female, female_pr, male, male_pr)

knitr::kable(tab1, caption = "Descriptive Table")
```

@@

Since some event numbers in "less than 15" group are too small, then we filter the visits which are no more than 20, and add them to "15-65" group, and set the events, OR, and CI are 0. (But the descriptive table above shows real data summary)

```{r echo = F, warning = F, results='asis'}
# data wrangling ----
# Producing conditional logit model estimates loop 
for(i in 1:length(df_list)){

@@

  # dataframe for analysis creation
  # bind columns back together 
  df_analysis <- covariates_df %>% 
    # remove missing pm values
    filter(!is.na(geo_smk_pm_zip10)) %>% 
    mutate(obs_diff_admission = (fromdate - date)/7) 
  # dataframe is already for the entire fire season, so I don't need to subset anymore

@@

@@

# combine each outcome dataframe itteration in to a big dataset
ed_results <- bind_rows(datalist)

knitr::kable(ed_results, caption = "GWR and ED admissions results")
```

@@

```{r time stratified age strata cdc met, warning =F, echo = F, results='asis'} 
# age category list
age_cat_list <- c(0,1,2)

@@

var_list <- c('respiratory', 'asthma', 'copd', 
              'cvd', 'isch_heart_dis',  'heart_failure', 
              'cerbrovas_dis')

@@

  # bind rows of age category estimates together
 age_est_df <- as.data.frame(bind_rows(datalist1))
 age_est_df[,5] <- as.numeric(age_est_df[,5])

 if(age_est_df[1,5] <= 20){
   age_est_df[2,5] <- age_est_df[2,5] + age_est_df[1,5]
   age_est_df[1,5] <- 0
   age_est_df[1,6] <- 0
   age_est_df[1,7] <- 0
   age_est_df[1,8] <- 0
 }

@@

# combine each outcome dataframe itteration in to a big dataset
ed_age_results <- bind_rows(datalist2) %>% 
  mutate(age_cat2 = ifelse(age_cat == 0, "Less than 15", 
                    ifelse(age_cat == 1, "15-65",
                    ifelse(age_cat == 2, "Greater than 65", NA)))) %>% 
  # subset columns I want to put in to the table
  mutate(odds_ratio = ifelse(odds_ratio == 99, "--", odds_ratio),
         lower95 = ifelse(lower95 == 99, "--", lower95),
         upper95 = ifelse(upper95 == 99, "--", upper95)) %>% 
  filter(n_events >= 50)

knitr::kable(ed_age_results, caption = "Age Stratified Results")
```

@@

```{r sex strata cdc met adj, warning = F, echo = F, results='asis'} 
# sex category list
sex_strata_list <- c(0,1)

@@

# combine each outcome dataframe itteration in to a big dataset
ed_sex_results <- bind_rows(datalist2) %>% 
  mutate(sex_cat = ifelse(sex == 0, "Male", 
               ifelse(sex == 1, "Female", NA))) %>%
  # subset columns I want to put in to the table
  select(1, 3, 11, 5:8)

# table
knitr::kable(ed_sex_results, caption = "Sex-stratified results")
```

@@

# print plot
ggsave(filename = "./r_scripts/analysis/Figs/cardiopulmonary_er_result.tiff", 
       plot = result_plot, 
       units="in", width=8, height=6, dpi=300)

@@

We see that the exposure method is significantly associated with an increase in asthma with Geo-Weighted Regression, those are associated with COPD. I believe this is the most justifiable design approach that would average any extreme values of PM~2.5~ across the wildfire season for the vector of referent values for a particular subject.
